---
title: "APG Contig Functional Characterization Plots"
author: "Abdulfatai Tijjani"
date: "2026-01-24"
output: html_document
---

This R Markdown file contains the plotting code used to generate Figures 5a-d in the APG contig manuscript.

```{r setup, message=FALSE, warning=FALSE}
# Core packages used across the Figure 5 workflow
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(tibble)
library(scales)
library(forcats)
library(cowplot)
```

# Global paths

```{r paths}
# Centralize input/output paths so the document is easier to move across systems.
# Update these paths only if the repository structure changes.

btc_data_dir    <- "../data/btc_functional"
figure_out_dir  <- file.path(btc_data_dir, "figures")

# Create the figure output directory if it does not already exist.
dir.create(figure_out_dir, recursive = TRUE, showWarnings = FALSE)
```

# Figure 5A. Distribution of APG contigs across alignment classes

```{r fig5A-data}
# Contig counts in the three manuscript alignment classes.
# These values are hard-coded because they represent final counts used in the figure.
fig5a_df <- tibble(
  category = c("T2T + HPRC", "HPRC only", "Below threshold"),
  count    = c(14796, 5649, 742)
)

fig5a_colors <- c(
  "T2T + HPRC"      = "steelblue",
  "HPRC only"       = "#E6A85C",
  "Below threshold" = "lightgrey"
)

fig5a_plot_df <- fig5a_df %>%
  mutate(
    category = factor(category, levels = names(fig5a_colors)),
    prop = count / sum(count),
    pct  = prop * 100,
    label = sprintf("%d (%.1f%%)", count, pct)
  )

fig5a_plot_df
```

```{r fig5A-plot, fig.width=4.8, fig.height=4.8}
# Single stacked bar summarizing the fraction of APG contigs in each class.
fig5A <- ggplot(fig5a_plot_df, aes(x = 1, y = prop, fill = category)) +
  geom_col(width = 0.55, color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 3.8
  ) +
  scale_fill_manual(
    values = fig5a_colors,
    drop = FALSE,
    name = "Reasonably Good \nAlignment"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_x_continuous(
    limits = c(0.7, 1.4),
    breaks = 1,
    labels = NULL,
    expand = c(0, 0)
  ) +
  labs(x = "APG contigs", y = "Fraction of contigs") +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.83, 0.25),
    legend.title = element_text(face = "bold"),
    legend.background = element_rect(fill = alpha("white", 0.85), color = "grey80"),
    plot.margin = margin(6, 6, 6, 6)
  )

fig5A
```

```{r fig5A-save, eval=FALSE}
# Optional export used for manuscript figure assembly.
ggsave(
  filename = file.path(figure_out_dir, "fig5A_alignment_class_distribution.png"),
  plot = fig5A,
  width = 4.8,
  height = 4.8,
  dpi = 300
)
```

# Figure 5B / 5C
# Contig evidence classes from predicted genes and CpG islands

## Load contig-level evidence tables

```{r evidence-data}
# Input tables:
# - btc_gene_functional_evidence.tsv: gene-level evidence table
# - phase4_cpg_contig_summary.tsv: contig-level CpG summary
# - btc_fa2_contig_lengths.tsv: contig lengths
# - btc_contig_functional_summary.tsv: consolidated contig-level summary
#
# Note:
# The consolidated summary table is used for the final merge because it already
# contains the key fields needed for Figure 5:
#   - n_genes
#   - n_genes_with_any_protein_support
#   - n_cpg_islands
#   - cpg_bp_total

gene_evidence_file        <- file.path(btc_data_dir, "btc_gene_functional_evidence.tsv")
cpg_summary_file          <- file.path(btc_data_dir, "phase4_cpg_contig_summary.tsv")
contig_length_file        <- file.path(btc_data_dir, "btc_fa2_contig_lengths.tsv")
functional_summary_file   <- file.path(btc_data_dir, "btc_contig_functional_summary.tsv")

# Check that all required input files are present before proceeding.
stopifnot(file.exists(gene_evidence_file))
stopifnot(file.exists(cpg_summary_file))
stopifnot(file.exists(contig_length_file))
stopifnot(file.exists(functional_summary_file))

# Optional upstream tables retained here for transparency/documentation.
# They are not required for the final merge below because the consolidated
# functional summary already contains the fields used in the figure.
contig_gene_ev <- read_tsv(gene_evidence_file, show_col_types = FALSE) %>%
  group_by(contig) %>%
  summarise(
    n_genes = n_distinct(gene_id),
    .groups = "drop"
  )

cpg <- read_tsv(cpg_summary_file, show_col_types = FALSE)

# Contig lengths are used as the base table so all BTC contigs are retained.
len <- read_tsv(
  contig_length_file,
  col_names = c("contig", "contig_length_bp"),
  show_col_types = FALSE
)

# Read the consolidated contig-level functional summary.
contig_functional_summary <- read_tsv(
  functional_summary_file,
  show_col_types = FALSE
)

# Confirm that the required columns are present in the summary table.
required_cols <- c(
  "contig",
  "n_genes",
  "n_genes_with_any_protein_support",
  "n_cpg_islands",
  "cpg_bp_total"
)

missing_cols <- setdiff(required_cols, colnames(contig_functional_summary))
if (length(missing_cols) > 0) {
  stop(
    "Missing required columns in btc_contig_functional_summary.tsv: ",
    paste(missing_cols, collapse = ", ")
  )
}

# Merge contig lengths with the consolidated functional summary.
# Missing values are set to zero so contigs without a given type of evidence
# are classified consistently in downstream plots.
final <- len %>%
  left_join(
    contig_functional_summary %>%
      select(
        contig,
        n_genes,
        n_genes_with_any_protein_support,
        n_cpg_islands,
        cpg_bp_total
      ),
    by = "contig"
  ) %>%
  mutate(
    n_genes = replace_na(n_genes, 0L),
    n_genes_with_any_protein_support = replace_na(n_genes_with_any_protein_support, 0L),
    n_cpg_islands = replace_na(n_cpg_islands, 0L),
    cpg_bp_total = replace_na(cpg_bp_total, 0)
  ) %>%
  mutate(
    evidence_class = case_when(
      n_genes > 0 & n_genes_with_any_protein_support > 0 & n_cpg_islands > 0 ~ "supported_gene_and_cpg",
      n_genes > 0 & n_genes_with_any_protein_support > 0 & n_cpg_islands == 0 ~ "supported_gene_only",
      n_genes > 0 & n_genes_with_any_protein_support == 0 & n_cpg_islands > 0 ~ "unsupported_gene_and_cpg",
      n_genes > 0 & n_genes_with_any_protein_support == 0 & n_cpg_islands == 0 ~ "unsupported_gene_only",
      n_genes == 0 & n_cpg_islands > 0 ~ "cpg_only",
      TRUE ~ "no_evidence"
    )
  )

# Quick check of evidence-class counts before collapsing to broader categories.
table(final$evidence_class)
```

```{r evidence-summary}
# Overall totals used for manuscript interpretation and QC.
totals <- final %>%
  summarise(
    n_contigs = n(),
    total_predicted_genes = sum(n_genes, na.rm = TRUE),
    total_cpg_islands = sum(n_cpg_islands, na.rm = TRUE),
    total_cpg_bp = sum(cpg_bp_total, na.rm = TRUE),
    mean_cpg_length_overall = ifelse(
      sum(n_cpg_islands, na.rm = TRUE) > 0,
      sum(cpg_bp_total, na.rm = TRUE) / sum(n_cpg_islands, na.rm = TRUE),
      NA_real_
    )
  )

totals
```

```{r evidence-4cat-summary}
# Collapse the six evidence classes into the four manuscript categories:
#   Both, Gene only, CpG Islands only, Neither
plot_df <- final %>%
  mutate(
    category = case_when(
      evidence_class %in% c("supported_gene_and_cpg", "unsupported_gene_and_cpg") ~ "Both",
      evidence_class %in% c("supported_gene_only", "unsupported_gene_only") ~ "Gene only",
      evidence_class == "cpg_only" ~ "CpG Islands only",
      evidence_class == "no_evidence" ~ "Neither",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(category)) %>%
  count(category, name = "n") %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, " (", percent(pct, accuracy = 0.1), ")"),
    category = factor(category, levels = c("Both", "Gene only", "CpG Islands only", "Neither"))
  ) %>%
  arrange(category) %>%
  mutate(ypos = cumsum(pct) - 0.5 * pct)

plot_df
```

## Figure 5C. Donut plot of four evidence classes

```{r fig5C-data}
# The final manuscript donut plot uses the fixed category counts below.
# Keeping these values explicit preserves the original figure exactly.
donut_df <- tribble(
  ~evidence_4cat,        ~n,
  "Both",               297,
  "Gene only",          149,
  "CpG Islands only",    27,
  "Neither",            269
)

donut_colors <- c(
  "Both"             = "#E7298A",
  "Gene only"        = "#3182BD",
  "CpG Islands only" = "#1B9E77",
  "Neither"          = "#BDBDBD"
)

plot_df2 <- donut_df %>%
  mutate(evidence_4cat = factor(evidence_4cat, levels = names(donut_colors))) %>%
  arrange(evidence_4cat) %>%
  mutate(
    pct = n / sum(n),
    label = paste0(n, "\n", percent(pct, accuracy = 0.1)),
    y_center = cumsum(n) - n / 2
  )

# Separate the tiny CpG slice for custom label placement.
df_main <- plot_df2 %>% filter(evidence_4cat != "CpG Islands only")
df_cpg  <- plot_df2 %>%
  filter(evidence_4cat == "CpG Islands only") %>%
  mutate(
    x_label = 2.35,
    x_line_end = 2.15
  )

plot_df2
```

```{r fig5C-plot, fig.width=7, fig.height=5}
fig5C <- ggplot(plot_df2, aes(x = 2, y = n, fill = evidence_4cat)) +
  geom_col(width = 1, color = "white", linewidth = 0.7) +
  geom_text(
    data = df_main,
    aes(label = label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4.5,
    lineheight = 0.95,
    show.legend = FALSE
  ) +
  geom_text(
    data = df_cpg,
    aes(x = 2.10, y = y_center - 170, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    color = "white",
    fontface = "bold",
    size = 4.2,
    lineheight = 0.95
  ) +
  coord_polar(theta = "y") +
  xlim(0.7, 2.6) +
  scale_fill_manual(values = donut_colors, name = NULL) +
  theme_void() +
  theme(
    legend.position = c(0.35, 0.5),
    legend.justification = c(0, 0.5),
    legend.key.size = unit(0.55, "lines"),
    legend.text = element_text(size = 12)
  )

fig5C
```

```{r fig5C-save, eval=FALSE}
ggsave(
  filename = file.path(figure_out_dir, "fig5C_donut_labels_legend_mid_final.png"),
  plot = fig5C,
  width = 7,
  height = 5,
  dpi = 300
)
```

# Figure 5D. Protein support classes among predicted genes

```{r fig5D-data}
# Load the functional gene evidence table and convert support columns to logicals.
func_gene_file <- file.path(btc_data_dir, "btc_gene_functional_evidence.tsv")

if (!file.exists(func_gene_file)) {
  stop(paste("Error: File not found at", func_gene_file, ". Please check the path."))
}

fg <- read_tsv(func_gene_file, show_col_types = FALSE)

fg2 <- fg %>%
  mutate(
    pfam_supported   = as.character(pfam_supported),
    blastp_supported = as.character(blastp_supported),
    pfam_supported   = pfam_supported %in% c("True", "TRUE", "T", "1", "TRUE ", "true"),
    blastp_supported = blastp_supported %in% c("True", "TRUE", "T", "1", "TRUE ", "true")
  ) %>%
  mutate(
    protein_class = case_when(
      pfam_supported & blastp_supported ~ "Pfam + BLASTP",
      pfam_supported & !blastp_supported ~ "Pfam only",
      !pfam_supported & blastp_supported ~ "BLASTP only",
      TRUE ~ "No support"
    ),
    protein_class = factor(
      protein_class,
      levels = c("Pfam + BLASTP", "Pfam only", "BLASTP only", "No support")
    )
  )

gene_protein_frac <- fg2 %>%
  count(protein_class) %>%
  mutate(frac = n / sum(n))

gene_protein_frac
```

```{r fig5D-plot, fig.width=7, fig.height=4.5}
# Horizontal bar plot of protein support categories.
fig5D_colors <- c(
  "Pfam + BLASTP" = "#CD661D",
  "Pfam only"     = "hotpink",
  "BLASTP only"   = "#A2CD5A",
  "No support"    = "#CDAA7D"
)

fig5D <- ggplot(gene_protein_frac, aes(x = protein_class, y = frac)) +
  geom_col(aes(fill = protein_class), color = "black", linewidth = 0.3) +
  coord_flip(clip = "off") +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 0.4),
    expand = expansion(mult = c(0, 0.08))
  ) +
  scale_fill_manual(values = fig5D_colors, guide = "none") +
  geom_text(
    aes(label = n),
    hjust = -0.1,
    size = 4,
    color = "black",
    family = "Arial"
  ) +
  labs(
    x = NULL,
    y = "Fraction of Augustus Predicted Genes (%)"
  ) +
  theme_cowplot(font_size = 12, font_family = "Arial") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16, margin = margin(b = 10)),
    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(b = 15)),
    plot.caption = element_text(hjust = 1, size = 9, color = "gray50", margin = margin(t = 15)),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 11, color = "black", face = "bold"),
    axis.title.x = element_text(size = 12, margin = margin(t = 10)),
    axis.line.x = element_line(color = "black", linewidth = 0.5),
    axis.line.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray85", linetype = "dotted"),
    plot.margin = margin(t = 15, r = 15, b = 15, l = 15),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7)
  )

fig5D
```

```{r fig5D-save, eval=FALSE}
ggsave(
  filename = file.path(figure_out_dir, "fig5D_protein_support_revised_color_final.png"),
  plot = fig5D,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

# Session information

```{r session-info}
sessionInfo()
```
