---
title: "Code to Generate Fig.5e - BTC Gene Expression Across Populations"
output: html_document
date: "2025-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data
Libraries
```{r}
library(readr)
library(dplyr)
library(data.table)
library(readxl)
library(ggplot2)
library(ComplexHeatmap)
library(tibble)
```
Load file with counts of samples per population group with BTC gene expression - 1000 Genomes cohort
```{r}
p1kg <- read.csv("1KGP_PopCounts_per_BTCgene.csv")
p1kg$cohort <- "1000 Genomes"
```

Load file with counts of samples per population group with BTC gene expression - TCGA BRCRA cohort
```{r}
tcga <- read.csv("TCGA_PopCounts_per_BTCgene.csv")
tcga$cohort <- "TCGA BRCA"
```

Identify BTC genes expressed (TPM value > 0) across any of the cohorts
```{r}
genes <- rbind(p1kg[,-3], tcga)
head(genes)
```

Load functional annotation support for BTC genes
```{r}
functional_support <- read_delim("btc_contig_level_functional_summary.tsv", 
     delim = "\t", escape_double = FALSE, 
     trim_ws = TRUE)

colnames(functional_support)[1] <- "accession"
```

Create dataframe with functional annotation support for expressed BTC genes across cohorts and populations
```{r}
merged <- merge(genes, functional_support[,-c(4,7,9,10,12)], by="accession")
merged <- merged[order(merged$AFR, merged$EUR, decreasing = T),]

df <- merged %>%
  mutate(
    support = case_when(
      n_genes_pfam_supported > 0 & n_genes_blastp_supported > 0 ~ "Pfam & BLAST",
      n_genes_pfam_supported > 0 ~ "Pfam only",
      n_genes_blastp_supported > 0 ~ "BLAST only",
      TRUE ~ "No support"
    )
  )

df$support <- factor(df$support, levels=c("No support", "BLAST only", "Pfam only", "Pfam & BLAST"))
head(df)
```

Reformat dataframe to be used for plotting
```{r}
library(tidyr)

df_long <- df %>%
  pivot_longer(c(AFR, EUR), names_to = "population", values_to = "n_assemblies")
```

## Generate visualization
Generate plot for all genes expressed across cohorts
```{r}
p1 <- ggplot(df_long, 
       aes(x = reorder(gene, n_assemblies),
           y = n_assemblies,
           fill = population)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~cohort) +
  coord_flip() +
  theme(
    text = element_text(size = 15),
    plot.title = element_text(size = 15, face = "bold"),
    axis.text.x = element_text(color = "black", hjust = 1,size = 15),
    axis.text.y = element_text(color = "black",size = 13),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(color = "black"),
    panel.border = element_rect(colour = "black", fill = NA, size = 1)
  )  +
  labs(
    x = "Predicted Gene",
    y = "Number of Assemblies with RNA Expression",
    fill = "Superpop"
  )+ 
  scale_fill_manual(values=c("darkorange","royalblue"))

p1
```

Save plot
```{r}
ggsave("all_expressed_btc_genes.jpeg",
       width = 18, height = 19, device='jpeg', dpi=700,units = c("cm"))
```

Generate order that the genes appear on the full expressed gene plot (p1)
```{r}
built <- ggplot_build(p2)
gene_order <- built$layout$panel_params[[1]]$y$get_labels()
```

Create plot for only top 12 genes
```{r}
p2 <- ggplot(df_long[which(df_long$gene %in% rev(gene_order)[1:12]),], 
       aes(x = reorder(gene, n_assemblies),
           y = n_assemblies,
           fill = population)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~cohort) +
  coord_flip() +
  theme(
    text = element_text(size = 15),
    plot.title = element_text(size = 15, face = "bold"),
    axis.text.x = element_text(color = "black", hjust = 1,size = 15),
    axis.text.y = element_text(color = "black",size = 13),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(color = "black"),
    panel.border = element_rect(colour = "black", fill = NA, size = 1)
  )  +
  labs(
    x = "Predicted Gene",
    y = "Number of Assemblies with RNA Expression",
    fill = "Superpop"
  )+ 
  scale_fill_manual(values=c("darkorange","royalblue"))

p2
```
Save plot with top BTC 12 genes
```{r}
ggsave("top-12_expressed_btc_genes.jpeg",
       width = 16, height = 12, device='jpeg', dpi=700,units = c("cm"))
```

Regenerate order of genes as they appear in the top-12-gene plot and save it as .csv file (to be used for the Martini et al. cohort visualization)
```{r}
support_gene <- df_long[which(df_long$gene %in% rev(gene_order)[1:12]),] %>%
  select(gene, support) %>%
  distinct()

support_vec <- factor(
  support_gene$support,
  levels = c("No support", "BLAST only", "Pfam only", "Pfam & BLAST")
)

names(support_vec) <- support_gene$gene

support_vec <- support_vec[rev(gene_order)[1:12]]

write.csv(names(support_vec), "gene_order_functional_plot.csv",
          row.names = F)
```

*Same code to be used to generate the Martini et al. cohort related panel fig.5e, but this time using ethnicity instead of superpopulation annotations per sample.*
