---
title: "Violin plots for contigs with differential coverage"
author: "Kyriaki Founta"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries & data
Libraries
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
```

Percent contig length coverage .csv file for all three cohorts was generated using the "HPRCcontigs_calculate_pct_contig_coverage.ipynb" script. Additionally, for every cohort, we added "superpop" (population or ethnicity group) annotation per sample as a new column in the .csv file, using the available metadata. 

1kgp percent contig length coverage
```{r}
dat1 <- read_csv("hprc_contig_pct_coverage_1kgp-superpop.csv")

dat1$Cohort <- "1000 Genomes"

head(dat1)
```

TCGA percent contig length coverage
```{r}
dat2 <- read_csv("hprc_contig_pct_coverage_tcga-superpop.csv")

dat2$Cohort <- "TCGA BRCA"

head(dat2)
```

Martini et al. percent contig length coverage
```{r}
dat3 <- read_csv("hprc_contig_pct_coverage_martini-superpop.csv")

dat3$Cohort <- "Martini et al."
colnames(dat3)[4] <- "superpop"

head(dat3)
```

Select contigs of interest (differentially RNAseq covered apg-contigs between AFR & EUR samples)
```{r}
contigs <- c("PDBU01048398.1", "PDBU01024832.1", "PDBU01001207.1", "PDBU01014857.1", "PDBU01050395.1", "PDBU01096400.1", "PDBU01000224.1", "PDBU01000303.1", "PDBU01001035.1")
```

Combine all cohort data
```{r}
all <- rbind(dat1[which(dat1$contig %in% contigs),], dat2[which(dat2$contig %in% contigs),], dat3[which(dat3$contig %in% contigs),])
head(all)
```

```{r}
all[which(all$superpop == "White"),"superpop"] <- "European American"
```


## Visualization 
Generate plot list for all contigs
```{r}
library(tidyverse)          
library(ggbeeswarm)        
library(patchwork)  
library(ggsignif)

# Define colors once      
my_colors <- c(      
  "AFR" = "darkorange",       
  "EUR" = "royalblue",      
  "African American" = "#8968CD",      
  "Ethiopian" = "#CD9B9B",      
  "Ghanaian" = "#6B8E23",      
  "European American" = "gray50"      
)

# Define which strand(s) to use for each contig    
strand_map <- list(      
  "PDBU01096400.1" = "str2",      
  "PDBU01000224.1" = "str2",      
  "PDBU01001035.1" = "str2",      
  "PDBU01000303.1" = "str2",      
  "PDBU01024832.1" = c("str1", "str2"),    
  "PDBU01014857.1" = "str2",      
  "PDBU01001207.1" = "str2",      
  "PDBU01048398.1" = "str1",      
  "PDBU01050395.1" = "str2"    
)

# ============================================================    
# PRE-COMPUTED SIGNIFICANCE RESULTS    
# Only include significant results (no "ns" rows needed)  
# group1 and group2 = the two superpops being compared  
# ============================================================    
sig_df <- tribble(    
  ~contig,             ~strand, ~Cohort,          ~group1, ~group2, ~label,  
  "PDBU01096400.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*",  
  "PDBU01000224.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*",  
  "PDBU01001035.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*",  
  "PDBU01000303.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*",  
  "PDBU01024832.1",    "str1",  "1000 Genomes",   "AFR",   "EUR",  "*",  
  "PDBU01024832.1",    "str2",  "1000 Genomes",   "AFR",   "EUR",  "*",  
  "PDBU01014857.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*",  
  "PDBU01001207.1",    "str2",  "1000 Genomes",   "AFR",   "EUR",  "*",  
  "PDBU01048398.1",    "str1",  "1000 Genomes",   "AFR",   "EUR",  "*",  
  "PDBU01050395.1",    "str2",  "TCGA BRCA",      "AFR",   "EUR",  "*"  
)

# Get unique contigs          
contigs <- unique(all$contig)        
cohorts <- unique(all$Cohort)

# Initialize empty list to store plots          
plot_list <- list()    
plot_index <- 1

for (i in seq_along(contigs)) {          
            
  contig_name <- contigs[i]          
      
  strands <- strand_map[[contig_name]]    
  if (is.null(strands)) strands <- "str1"    
      
  for (strand in strands) {    
        
    coverage_col <- paste0("pct_covered_T100_", strand)      
    strand_label <- ifelse(strand == "str1", "Strand 1", "Strand 2")    
        
    df_contig <- all %>%          
      dplyr::filter(contig == contig_name) %>%      
      mutate(coverage_value = .data[[coverage_col]])    
            
    cohort_plots <- list()        
            
    for (j in seq_along(cohorts)) {        
              
      cohort_name <- cohorts[j]        
              
      df_subset <- df_contig %>%          
        dplyr::filter(Cohort == cohort_name) %>%          
        mutate(superpop = factor(superpop,   
                                 levels = c("AFR", "EUR",   
                                            "African American", "Ethiopian",   
                                            "Ghanaian", "European American")))      
              
      if (nrow(df_subset) == 0) next    
        
      # Look up if there's a significant result for this contig + strand + cohort  
      sig_subset <- sig_df %>%  
        dplyr::filter(contig == contig_name, strand == !!strand, Cohort == cohort_name)  
          
      p_cohort <- ggplot(df_subset, aes(x = superpop, y = coverage_value, fill = superpop)) +          
        geom_violin(alpha = 0.5, outlier.shape = NA) +  
        stat_summary(fun = mean, geom = "crossbar",  width = 0.15,  
                     fatten = 1.5, color = "black", linewidth = 0.6,   
                     fill = "black", color = "black") +  
        #geom_boxplot(alpha = 0.5, outlier.shape = NA) +          
        geom_beeswarm(aes(color = superpop), size = 1.5, alpha = 0.7, cex = 2) +          
        scale_fill_manual(values = my_colors) +      
        scale_color_manual(values = my_colors) +      
        scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +    
        labs(          
          title = cohort_name,        
          x = NULL,          
          y = NULL        
        ) +          
        theme(          
          text = element_text(size = 12),
          plot.title = element_text(hjust = 0.5, face = "bold", size = 12),          
          axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),  
          axis.text.y = element_text(color = "black"),  
          legend.position = "none",  
          panel.background = element_blank(),    
          axis.line = element_line(colour = "black"),    
          panel.border = element_rect(colour = "black", fill = NA, size = 1)  
        )  
        
      # Add significance bracket only if significant  
      if (nrow(sig_subset) > 0) {  
        y_max <- max(df_subset$coverage_value, na.rm = TRUE)  
          
        for (k in 1:nrow(sig_subset)) {  
          p_cohort <- p_cohort +  
            geom_signif(  
              comparisons = list(c(sig_subset$group1[k], sig_subset$group2[k])),  
              annotations = sig_subset$label[k],  
              y_position = y_max * 1.05 + (y_max * 0.1 * (k - 1)),  
              tip_length = 0.02,  
              textsize = 5,  
              color = "black"  
            )  
        }  
          
        # Extend y-axis to fit brackets  
        p_cohort <- p_cohort +  
          scale_y_continuous(  
            limits = c(0, y_max * 1.25),  
            expand = expansion(mult = c(0, 0.05))  
          )  
      }  
            
      cohort_plots[[j]] <- p_cohort        
    }        
            
    cohort_plots <- compact(cohort_plots)        
            
    combined <- wrap_plots(cohort_plots, nrow = 1) +        
      plot_annotation(        
        title = paste("Coverage at 100x -", contig_name, paste0("(", strand_label, ")")),        
        subtitle = paste("Contig length:", unique(df_contig$length), "bp"),        
        theme = theme(        
          plot.title = element_text(hjust = 0.5, face = "bold", size = 14),        
          plot.subtitle = element_text(hjust = 0.5, size = 10)        
        )        
      )        
            
    combined <- combined &        
      ylab("% Coverage at 100x") &        
      xlab("")    
        
    plot_list[[plot_index]] <- combined    
    plot_index <- plot_index + 1    
  }    
}

plot_list  
```

Save plots
```{r}
# Save all plots into a single PDF  
pdf("violin_beeswarm_boxplots_100x_AllCohorts_sig_contigs.pdf", width = 10, height = 5)  
for (p in plot_list) {  
  print(p)  
}  
dev.off()

plot_list  
```

