---
title: "APG Contig Alignment to HPRC Assemblies"
author: "Abdulfatai Tijjani"
date: "2025-11-20"
output: html_document
---

This R Markdown document reproduces the APG manuscript analyses and plots (Figure 4b_c and Supplementary Figures 5a_b) related to contig alignment across HPRC superpopulations and individual HPRC linear assemblies.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(readxl)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(UpSetR)
  library(forestplot)
  library(grid)
  library(patchwork)
  #library(ComplexHeatmap)
  library(circlize)
})
```

# 1. Continental-level APG contig alignment across HPRC superpopulations

## 1.1 Load superpopulation-level alignment data

```{r load-superpopulation-data}
# Input: Supplementary Table ST12
# Expected columns include AFR, AMR, EAS, EUR, and SAS.
input_excel <- "../../../apg_contigs_paper/Supplemental_tables_S12-S16.xlsx"
input_sheet <- "ST12"

superpop_hprc <- read_excel(
  path = input_excel,
  sheet = input_sheet,
  skip = 1
)

# Validate that all required superpopulation columns are present.
required_cols <- c("AFR", "AMR", "EAS", "EUR", "SAS")
missing_cols <- setdiff(required_cols, colnames(superpop_hprc))

if (length(missing_cols) > 0) {
  stop(
    "Missing required columns in ST12: ",
    paste(missing_cols, collapse = ", ")
  )
}

# Keep a simple 0/1 binary matrix for downstream analyses and plotting.
upset_df <- superpop_hprc[, required_cols]
upset_df[] <- lapply(upset_df, function(x) as.numeric(trimws(as.character(x))))

head(superpop_hprc)
```

## 1.2 Summarize aligned and unaligned contigs per superpopulation

```{r summarize-superpopulation-counts}
# Number of HPRC samples per superpopulation used in this analysis.
pop_size <- c(
  AFR = 24,
  AMR = 16,
  EAS = 5,
  EUR = 1,
  SAS = 1
)

# Summarize contig alignment presence/absence within each superpopulation.
alignment_counts <- superpop_hprc %>%
  summarise(
    AFR_aligned = sum(AFR == 1),
    AFR_unaligned = sum(AFR == 0),
    AMR_aligned = sum(AMR == 1),
    AMR_unaligned = sum(AMR == 0),
    EAS_aligned = sum(EAS == 1),
    EAS_unaligned = sum(EAS == 0),
    EUR_aligned = sum(EUR == 1),
    EUR_unaligned = sum(EUR == 0),
    SAS_aligned = sum(SAS == 1),
    SAS_unaligned = sum(SAS == 0)
  )

alignment_counts
```

## 1.3 Normalize alignment counts by number of samples per superpopulation

```{r normalize-superpopulation-counts}
# Descriptive normalization: number of aligned contigs divided by the number of
# HPRC samples in each superpopulation.
normalized_alignment_rate <- c(
  AFR = alignment_counts$AFR_aligned / pop_size["AFR"],
  AMR = alignment_counts$AMR_aligned / pop_size["AMR"],
  EAS = alignment_counts$EAS_aligned / pop_size["EAS"],
  EUR = alignment_counts$EUR_aligned / pop_size["EUR"],
  SAS = alignment_counts$SAS_aligned / pop_size["SAS"]
)

normalized_alignment_rate
```

## 1.4 Global chi-square test across superpopulations

```{r chi-square-test}
# Build a 2 x 5 contingency matrix: aligned vs unaligned across superpopulations.
contingency_matrix <- matrix(
  c(
    alignment_counts$AFR_aligned, alignment_counts$AFR_unaligned,
    alignment_counts$AMR_aligned, alignment_counts$AMR_unaligned,
    alignment_counts$EAS_aligned, alignment_counts$EAS_unaligned,
    alignment_counts$EUR_aligned, alignment_counts$EUR_unaligned,
    alignment_counts$SAS_aligned, alignment_counts$SAS_unaligned
  ),
  nrow = 2,
  byrow = TRUE
)

colnames(contingency_matrix) <- c("AFR", "AMR", "EAS", "EUR", "SAS")
rownames(contingency_matrix) <- c("Aligned", "Not_Aligned")

contingency_matrix

# Global chi-square test for differences in aligned vs unaligned counts.
chi_result <- chisq.test(contingency_matrix)
chi_result
```

## 1.5 Pairwise odds ratios comparing AFR with other superpopulations

```{r odds-ratios}
# Compute simple odds ratios using aligned vs unaligned counts.
# This preserves the original manuscript logic used for the forest plot.
compute_odds_ratio <- function(aligned1, unaligned1, aligned2, unaligned2) {
  odds1 <- aligned1 / unaligned1
  odds2 <- aligned2 / unaligned2
  OR <- odds1 / odds2
  return(OR)
}

odds_AFR_vs_AMR <- compute_odds_ratio(
  alignment_counts$AFR_aligned, alignment_counts$AFR_unaligned,
  alignment_counts$AMR_aligned, alignment_counts$AMR_unaligned
)

odds_AFR_vs_EAS <- compute_odds_ratio(
  alignment_counts$AFR_aligned, alignment_counts$AFR_unaligned,
  alignment_counts$EAS_aligned, alignment_counts$EAS_unaligned
)

odds_AFR_vs_EUR <- compute_odds_ratio(
  alignment_counts$AFR_aligned, alignment_counts$AFR_unaligned,
  alignment_counts$EUR_aligned, alignment_counts$EUR_unaligned
)

odds_AFR_vs_SAS <- compute_odds_ratio(
  alignment_counts$AFR_aligned, alignment_counts$AFR_unaligned,
  alignment_counts$SAS_aligned, alignment_counts$SAS_unaligned
)

odds_ratios <- data.frame(
  Comparison = c("AFR_vs_AMR", "AFR_vs_EAS", "AFR_vs_EUR", "AFR_vs_SAS"),
  Odds_Ratio = c(odds_AFR_vs_AMR, odds_AFR_vs_EAS, odds_AFR_vs_EUR, odds_AFR_vs_SAS)
)

odds_ratios
```

## 1.6 Forest plot of odds ratios

```{r forest-plot}
# APG Manuscript Figure 4B
# Odds ratios comparing the likelihood of APG contig alignment in AFR genomes
# versus other superpopulations.

tabletext <- cbind(
  c("Comparison", "AFR vs AMR", "AFR vs EAS", "AFR vs EUR", "AFR vs SAS"),
  c("Odds Ratio", round(odds_ratios$Odds_Ratio, 2))
)

forestplot(
  labeltext = tabletext,
  mean = c(NA, odds_ratios$Odds_Ratio),
  lower = c(NA, odds_ratios$Odds_Ratio * 0.8),
  upper = c(NA, odds_ratios$Odds_Ratio * 1.2),
  zero = 1,
  boxsize = 0.2,
  colgap = unit(5, "mm"),
  xlab = "Odds Ratio (AFR vs Other Populations)",
  col = fpColors(box = "darkblue", line = "gray", summary = "royalblue")
)
```

## 1.7 Pattern heatmap of binary superpopulation sharing

```{r define-superpopulation-pattern-heatmap}
# APG Manuscript Supplementary Figure 5A
# Continental-level binary alignment patterns across HPRC linear assemblies,
# where a contig is aligned to at least one member of each of the five human
# superpopulations.
plot_binary_patterns_complexheatmap_transposed <- function(df, output_file = "pattern_heatmap_transposed.jpeg") {

  # Ensure numeric 0/1 values.
  df[] <- lapply(df, function(x) suppressWarnings(as.numeric(as.character(x))))

  # Preserve the original superpopulation column order.
  column_order <- colnames(df)

  # Collapse contigs into unique binary alignment patterns and count frequency.
  row_patterns <- df %>%
    group_by(across(all_of(column_order))) %>%
    summarise(Frequency = n(), .groups = "drop") %>%
    mutate(
      Aligned_Count = rowSums(across(all_of(column_order))),
      Alignment_Rank = 6 - Aligned_Count,
      Pattern_Key = apply(across(all_of(column_order)), 1, paste0, collapse = ""),
      Pattern_ID = paste0("P", row_number())
    ) %>%
    arrange(Alignment_Rank, desc(Frequency))

  # Matrix of unique patterns.
  pattern_matrix <- row_patterns %>%
    select(all_of(column_order)) %>%
    as.matrix()

  # Transpose so rows are superpopulations and columns are unique patterns.
  pattern_matrix_t <- t(pattern_matrix)

  # Enforce manuscript row order.
  desired_order <- c("AFR", "AMR", "EAS", "EUR", "SAS")
  desired_order <- desired_order[desired_order %in% rownames(pattern_matrix_t)]
  pattern_matrix_t <- pattern_matrix_t[desired_order, ]

  row_labels <- rownames(pattern_matrix_t)

  # Top annotation showing the number of contigs represented by each pattern.
  frequency_colors <- colorRamp2(
    c(min(row_patterns$Frequency), max(row_patterns$Frequency)),
    c("lightblue", "darkblue")
  )

  top_anno <- HeatmapAnnotation(
    Frequency = row_patterns$Frequency,
    col = list(Frequency = frequency_colors),
    which = "column",
    annotation_name_side = "left",
    annotation_legend_param = list(title = "Contig Count")
  )

  ht <- Heatmap(
    pattern_matrix_t,
    name = "Alignment Status",
    col = c("0" = "grey", "1" = "black"),
    heatmap_legend_param = list(
      title = "Alignment Status",
      at = c(0, 1),
      labels = c("Unaligned", "Aligned")
    ),
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    row_labels = row_labels,
    show_column_names = FALSE,
    column_title = "Contig Alignment Pattern",
    row_title = NULL,
    top_annotation = top_anno
  )

  return(list(summary = row_patterns, heatmap = ht))
}
```

```{r make-superpopulation-pattern-heatmap}
result <- plot_binary_patterns_complexheatmap_transposed(upset_df)
hprc_superpop_sum <- result$summary
hprc_superpop_ht <- result$heatmap
```

```{r}
hprc_superpop_ht
```

```{r export-superpopulation-pattern-heatmap, eval=FALSE}
# Export manuscript-ready image.
jpeg(
  "../plots/fig_S5A_apg_contigs_hprc_pop_alignment_new_final.jpeg",
  width = 2700,
  height = 700,
  units = "px",
  res = 180,
  bg = "white"
)
draw(hprc_superpop_ht)
dev.off()
```

## 1.8 UpSet plot of contig sharing across superpopulations

```{r superpopulation-upset-plot}
# APG Manuscript Supplementary Figure 5B
# UpSet plot depicting the number of intersections of contig alignments across
# human superpopulations.
library(ComplexUpset)

# Ensure integer 0/1 values for plotting.
upset_df[] <- lapply(upset_df, function(x) as.integer(as.numeric(as.character(x))))

# Reverse order so the display is AFR, AMR, EAS, EUR, SAS from top to bottom.
sets_order <- c("SAS", "EUR", "EAS", "AMR", "AFR")

plain_theme <- theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  text = element_text(size = 12)
)

upset_plot <- ComplexUpset::upset(
  upset_df,
  intersect = sets_order,
  name = NULL,
  sort_sets = FALSE,
  sort_intersections_by = "cardinality",
  set_sizes = upset_set_size(
    position = "right",
    geom = geom_bar(fill = "gray60", width = 0.6)
  ) +
    geom_text(
      aes(label = after_stat(count)),
      stat = "count",
      hjust = -0.2,
      size = 3
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
    ylab("Number of Contigs"),
  base_annotations = list(
    "Number of Contigs" = intersection_size(
      mapping = aes(fill = "steelblue"),
      text = list(
        angle = 45,
        hjust = -0.3,
        vjust = 0.2,
        size = 3
      )
    ) +
      scale_fill_manual(values = c("steelblue" = "steelblue"), guide = "none") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
      theme(
        axis.title.y = element_text(size = 11),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  ),
  matrix = intersection_matrix(
    geom = geom_point(size = 3),
    segment = geom_segment(linewidth = 1)
  ),
  themes = upset_modify_themes(list(
    "intersections_matrix" = plain_theme + theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    ),
    "overall_sizes" = plain_theme + theme(
      axis.title.x = element_text(size = 11)
    ),
    "Number of Contigs" = plain_theme,
    default = plain_theme
  )),
  width_ratio = 0.3,
  min_size = 0,
  wrap = FALSE
)

upset_plot
```

# 2. AFR-specific and AFR-AMR shared contigs across individual HPRC assemblies

## 2.1 Load summary and define contig groups of interest

```{r load-hprc-population-summary}
# Summary table used in the original workflow.
apg_hprc_pop_summary <- read.table("../results/unaligned_t2t_contig_hprc_pop_summary.txt")
head(apg_hprc_pop_summary)
```

```{r define-pattern-membership}
# Assign each contig to a binary superpopulation pattern.
contig_patterns <- superpop_hprc %>%
  group_by(across(-ContigID)) %>%
  mutate(Pattern_ID = paste0("P", cur_group_id())) %>%
  ungroup() %>%
  select(ContigID, Pattern_ID)

table(contig_patterns$Pattern_ID)
```

```{r define-afr-specific-contigs}
# P16 corresponds to AFR-specific contigs in this workflow.
afr_specific_contigs_P16 <- contig_patterns %>%
  filter(Pattern_ID == "P16") %>%
  mutate(Group = "AFR-specific") %>%
  dplyr::select(ContigID, Group) %>%
  distinct()

nrow(afr_specific_contigs_P16)
head(afr_specific_contigs_P16)
```

```{r define-afr-amr-shared-contigs}
# P24 corresponds to contigs shared between AFR and AMR in this workflow.
afr_amr_shared_contigs_P24 <- contig_patterns %>%
  filter(Pattern_ID == "P24") %>%
  mutate(Group = "AFR-AMR shared") %>%
  dplyr::select(ContigID, Group) %>%
  distinct()

nrow(afr_amr_shared_contigs_P24)
head(afr_amr_shared_contigs_P24)
```

```{r merge-contig-metadata}
# Combined contig metadata for the individual-level heatmap.
contig_metadata_afr_amr <- rbind(afr_specific_contigs_P16, afr_amr_shared_contigs_P24)

nrow(contig_metadata_afr_amr)
table(contig_metadata_afr_amr$Group)
head(contig_metadata_afr_amr)
```

## 2.2 Load sample metadata and individual alignment matrix

```{r load-sample-metadata}
# Sample metadata for HPRC individuals.
# Replace with the final curated metadata file if needed.
metadata <- read_excel(
  "../data/ST2_hprc_ancestry_annotations.xlsx",
  sheet = 1
) %>%
  dplyr::select(
    sampleID = Ind_name,
    superpopulation = Superpopulation_code
  )

head(metadata)
table(metadata$superpopulation)
dim(metadata)

metadata_afr <- metadata %>%
  group_by(superpopulation) %>%
  filter(superpopulation == "AFR")

metadata_amr <- metadata %>%
  group_by(superpopulation) %>%
  filter(superpopulation == "AMR")

metadata_afr_amr <- rbind(metadata_afr, metadata_amr) %>%
  rename(
    SampleID = sampleID,
    Superpopulation = superpopulation
  ) %>%
  as.data.frame()

nrow(metadata_afr_amr)
```

```{r load-individual-alignment-matrix}
# Individual HPRC alignment matrix.
# Replace this file path with the final curated Supplementary Table ST13 if needed.
hprc_ind_matrix <- read_csv("../results/unaligned_t2t_contig_hprc_ind47.csv") %>%
  select(ContigID, all_of(metadata_afr_amr$SampleID))

head(hprc_ind_matrix)
dim(hprc_ind_matrix)
```

```{r subset-afr-amr-alignment-matrix}
alignment_df_afr_amr <- hprc_ind_matrix %>%
  select(ContigID, all_of(metadata_afr_amr$SampleID))

dim(alignment_df_afr_amr)
dim(contig_metadata_afr_amr)
dim(metadata_afr_amr)
```

```{r assign-afr-amr-objects}
# Assign canonical objects used by the heatmap code below.
alignment_df <- alignment_df_afr_amr
contig_metadata <- contig_metadata_afr_amr
sample_metadata <- metadata_afr_amr
```

## 2.3 Individual-level heatmap across AFR and AMR assemblies

```{r build-afr-amr-individual-heatmap}
# APG Manuscript Figure 4C
# Heatmap showing APG contig alignment across individual linear assemblies from
# African (AFR) and Admixed American (AMR) HPRC samples.

# Step 1: Prepare the alignment matrix.
alignment_matrix <- alignment_df %>%
  tibble::column_to_rownames("ContigID") %>%
  as.matrix() %>%
  t()

# Step 2: Reorder sample metadata to match matrix row order exactly.
sample_metadata <- sample_metadata %>%
  filter(SampleID %in% rownames(alignment_matrix)) %>%
  arrange(match(SampleID, rownames(alignment_matrix)))

alignment_matrix <- alignment_matrix[sample_metadata$SampleID, ]

# Step 3: Reorder contigs so AFR-specific contigs appear first, followed by
# AFR-AMR shared contigs.
Contig_metadata <- contig_metadata_afr_amr %>%
  filter(ContigID %in% colnames(alignment_matrix)) %>%
  arrange(factor(Group, levels = c("AFR-specific", "AFR-AMR shared")))

alignment_matrix <- alignment_matrix[, Contig_metadata$ContigID]

# Step 4: Define color mappings.
alignment_colors <- c("1" = "black", "0" = "lightgray")
superpop_colors <- c("AFR" = "forestgreen", "AMR" = "steelblue")
group_colors <- c("AFR-specific" = "red", "AFR-AMR shared" = "purple")

# Step 5: Row annotation for sample superpopulation.
row_ha <- rowAnnotation(
  Superpopulation = sample_metadata$Superpopulation,
  col = list(Superpopulation = superpop_colors),
  show_annotation_name = FALSE,
  annotation_legend_param = list(title_gp = gpar(fontsize = 14))
)

# Step 6: Top annotation for contig group.
col_ha <- HeatmapAnnotation(
  Group = Contig_metadata$Group,
  col = list(Group = group_colors),
  show_annotation_name = TRUE,
  annotation_name_side = "right",
  annotation_legend_param = list(title_gp = gpar(fontsize = 14))
)

# Step 7: Build heatmap.
fig4C <- Heatmap(
  alignment_matrix,
  name = "Alignment",
  col = alignment_colors,
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  column_split = factor(
    Contig_metadata$Group,
    levels = c("AFR-AMR shared", "AFR-specific")
  ),
  column_gap = unit(2, "mm"),
  row_dend_side = "left",
  top_annotation = col_ha,
  left_annotation = row_ha,
  border = TRUE,
  heatmap_legend_param = list(
    title = "Alignment",
    at = c(1, 0),
    labels = c("Aligned", "Unaligned"),
    title_gp = gpar(fontsize = 14),
    labels_gp = gpar(fontsize = 14),
    direction = "horizontal"
  ),
  row_names_gp = gpar(fontsize = 14),
  column_title = "APG Contigs",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "HPRC Individual Linear Assemblies",
  row_title_gp = gpar(fontsize = 14, fontface = "bold")
)

#fig4C
```

```{r export-afr-amr-individual-heatmap, eval=FALSE}
# Export manuscript-ready image.
jpeg(
  "../plots/Fig4C_apg_contigs_hprc_afr_amr_spe_shared_combined.jpeg",
  width = 1000,
  height = 1500,
  res = 150
)
draw(
  fig4C,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "bottom",
  merge_legend = TRUE,
  newpage = TRUE
)
dev.off()
```
